An analysis of PCA image reconstruction for classification

### 2010-04-26 paper/paper.tex
Fitness function description on paper.
Adição do algoritmo, da conclusão e das referências.

The idea of principal component analysis (PCA) seems to have been first proposed by Pearson in 1901 \cite{pearson1901} and was later developed in many works during the last century, being now known as the Karhunen-Loève extension, the Hotelling transform or proper orthogonal decomposition, depending of the field of application \cite{serovich87}. It was popularized in computer vision by Turk and Pentland, that in 1991 used PCA to propose the eigenfaces \cite{turk91}, a face recognition method still relevant and used in current days. % The paper is currently the most cited work in computer vision.

@@

In the classification phase, when a new grayscale image has to be classified, the first step is obtain its edge image and perform four reconstructions: one for each set of PCs. Then, the reconstruction errors must be calculated and combined to produce the total error, that is the final classification score. If the total error is greater than or equal to zero, the image is classified as a pedestrian; otherwise, it is classifier as a non-pedestrian. The detailed procedure is described on Algorithm XX.

% TODO Algorithm description

@@

The non-pedestrian images were extracted from 120 assorted pictures that did not contain pedestrians. Each one was chopped into slices of $45 \times 105$ pixels and 5.000 slices were randomly chosen to compose the negative dataset. This dataset is more than four times larger than the positive dataset and that is related with the representation capacity of each one. The object detection problem is naturally asymmetric \cite{jiang09}, since the positive examples represents a specific type of object while the negative examples represent the \"rest of the world\". That is why we need much more negative examples than positive ones.

To train the classifier, we selected 75\% of the pedestrian images (693 images) and the remaining (231 images) was used  for testing; from the non-pedestrian images, we selected 80\% of the images (4.000 images) for training the classifier and the remaining (1.000 images) was used for testing. This particular scheme was chosen in a empirical way; it yielded the best results among the distributions we tested. The edge images are computed with $x$ and $y$ Sobel filters, which we combine to generate a single edge image. To implement the classifier and the auxiliary procedures, we used the MATLAB 2009b environment. The resulting code is available at \cite{code}.

@@

% What reference use to GA?
These findings also point out that, among all the four reconstruction errors, some are more relevant to the classification than others. However, on the total reconstruction error (equation \ref{total_error}) each error contributes equally. So, we use the weighted total reconstruction error (equation \ref{weighted_total_error}) to improve the classification accuracy. The problem with using weights is how to choose them. We can try some values and make adjusts manually, test after test. But we can also employ an optimization algorithm that looks for the values in an automated way. In this work, we use a genetic algorithm \ref{ga}. It initially chooses a set of random values and uses operations as crossover, mutation and selection to improve them during many iterations. These operations are inspired by evolutionary biology; hence the name genetic algorithms. The values are adjusted in relation to a fitness function. In our case, this functions aims to correctly classify the largest number of training samples using a given set of weights.

@@

>>>>

The idea of principal component analysis (PCA) seems to have been first proposed by Pearson in 1901 \cite{pearson1901} and was later developed in many works during the last century, being now known as the Karhunen-Loève extension, the Hotelling transform or proper orthogonal decomposition, depending of the field of application \cite{sirovich87}. It was popularized in computer vision by Turk and Pentland, that in 1991 used PCA to propose the eigenfaces \cite{turk91}, a face recognition method still relevant and used in current days. % The paper is currently the most cited work in computer vision.

@@

In the classification phase, when a new grayscale image has to be classified, the first step is obtain its edge image and perform four reconstructions: one for each set of PCs. Then, the reconstruction errors must be calculated and combined to produce the total error, that is the final classification score. If the total error is greater than or equal to zero, the image is classified as a pedestrian; otherwise, it is classifier as a non-pedestrian. The detailed procedure is described above.

\begin{enumerate}
  \item Obtain the edge image $e$ from grayscale image $g$ using the Sobel operator.
  \item Perform four image reconstructions:
    \begin{enumerate}
      \item $r_{gp} = P^T P_{gp}(g-u) + \mu_{gp}$
      \item $r_{ep} = P^T P_{ep}(e-u) + \mu_{ep}$
      \item $r_{gn} = P^T P_{gn}(g-u) + \mu_{gn}$
      \item $r_{en} = P^T P_{en}(e-u) + \mu_{en}$
    \end{enumerate}
  \item Calculate the corresponding reconstruction errors:
    \begin{enumerate}
      \item $d_{gp} = |r_{gp} - g|$
      \item $d_{ep} = |r_{ep} - e|$
      \item $d_{gn} = |r_{gp} - g|$
      \item $d_{en} = |r_{en} - e|$
    \end{enumerate}
  \item Compute the total reconstruction error, given by
    \begin{equation}
      d_t = d_{gn} + d_{en} - d_{gp} - d_{ep}
      \label{total_error}
    \end{equation}
  \item Classify the image according to:
    \begin{equation}
      \textrm{class}(g) = \left\{
        \begin{array}{l l}
         \textrm{Pedestrian}, & d_t >= 0\\
         \textrm{Non-pedestrian}, & d_t < 0
        \end{array}
      \right.
    \end{equation}

\end{enumerate}

@@

The non-pedestrian images were extracted from 120 assorted pictures that did not contain pedestrians. Each one was chopped into slices of $45 \times 105$ pixels and 5.000 slices were randomly chosen to compose the negative dataset. This dataset is more than four times larger than the positive dataset and that is related with the representation capacity of each one. The object detection problem is naturally asymmetric \cite{jiang09}, since the positive examples represents a specific type of object while the negative examples represent the ``rest of the world''. That is why we need much more negative examples than positive ones.

To train the classifier, we selected 75\% of the pedestrian images (693 samples) and the remaining images (231 samples) were used  for testing; from the non-pedestrian images, we selected 80\% of the images (4.000 samples) for training the classifier and the remaining images (1.000 samples) was used for testing. This particular scheme was chosen in a empirical way; it yielded the best results among the distributions we tested. The edge images are computed with $x$ and $y$ Sobel filters \cite{gonzalez01}, which we combine to generate a single edge image. To implement the classifier and the auxiliary procedures, we used the MATLAB 2009b environment. The resulting code is available at \cite{code}.

@@

% What reference use to GA?
These findings also point out that, among all the four reconstruction errors, some are more relevant to the classification than others. However, on the total reconstruction error (equation \ref{total_error}) each error contributes equally. So, we use the weighted total reconstruction error (equation \ref{weighted_total_error}) to improve the classification accuracy. The problem with using weights is how to choose them. We can try some values and make adjusts manually, test after test. But we can also employ an optimization algorithm that looks for the values in an automated way. In this work, we use a genetic algorithm \cite{ga}. It initially chooses a set of random values and uses operations as crossover, mutation and selection to improve them during many iterations. These operations are inspired by evolutionary biology; hence the name genetic algorithms. The values are adjusted in relation to a fitness function. In our case, this functions aims to correctly classify the largest number of training samples using a given set of weights.

@@

In this work we analyzed a classifier that uses PCA image reconstruction for pedestrian detection. It uses four sets of principal components (PCs): two originated from grayscale and edge images that contains pedestrians, respectively, and two originated from grayscale and edge images that did not contain pedestrians, respectively. To classify an image, it uses these sets of PCs to perform four image reconstructions and compares the respective reconstruction errors to determine if the given image depicts or not a pedestrian.

To understand how the individual errors contribute to the overall performance, we compared two classifiers: one that uses only the reconstructions from the positive PCs and other that uses only the reconstructions from the negative PCs. In our experiments, both performed poorly when compared to the full classifier, indicating that positives and negatives reconstructions are indeed relevant to the classification. The crucial role of the reconstructions from negative PCs is particularly surprising, since non-pedestrian samples are not related and do not contain a specific pattern. Therefore, we conclude that the importance of the negative reconstructions is not on what they characterize, but instead on what they do not characterize. In this way, the negative reconstruction errors contribute to raise the total reconstruction error when the image does not contains a pedestrian, what makes possible a better classification.

We also investigated the relevance of grayscale and edges errors. In other experiment we compared the accuracy of a classifier that uses only grayscale reconstructions and other classifier that uses only edge reconstructions. This time, they performed very well, with a performance close to the that of the total error classifier. This suggest that there is a lot of redundancy between grayscale and edge information. The classifier based only on edge images got better results, indicating that edge information is more relevant to this classifier than grayscale information. In fact, the edge classifier performed better than the total error classifier itself for some number of PCs and, when the results were worse, the difference between their accuracy was less than 1\%. For this reason, it is possible to speed up the classification time in a significant way by using only edge information and keep the classification accuracy virtually the same.

Since our experiments show that each error has a different relevance for classification, we proposed a classifier that uses a weighted version of the total reconstruction error. The problem with weights, though, is how to find the best values. To get an approximation in an automated way, we used a genetic algorithm with a fitness function that aims the best classification rates for the training samples. This classified performed better than the original version in our experiments for all numbers of PCs we tested. % DESCRIBE CLASSIFICATION RATE VS K

Future work include testing this classifier with other pedestrian databases and verifying if the reconstruction approach is suitable to other detection problems in computer vision, such as face detection. We intend to develop the classifier into a complete detection system, capable of locate objects in large pictures. It is also possible to employ variations of PCA for reconstruction, like the more robust 2D-PCA \cite{li05}.

% Mover para a introdução
% It classifies based on how much a reconstruction of an image is different from the original one, that is, it classifies based on the image reconstruction error. This relies on the fact that PCA aims to minimize the least mean square error for a given set of samples and therefore the reconstruction errors for images similar to the samples are lesser than the reconstruction errors for dissimilar images. So, we train the classifier by doing PCA separately for pedestrians and non-pedestrians. When an image has to be classified, we compare the reconstructions made by both sets of principal components to determine if it contains or not a pedestrian.
